Zero-shot task adaptation by homoiconic meta-mapping

### 2019-10-29 writing/homoiconic_meta_mapping_main.tex
Preprint version
Tweaks and formatting for camera-ready for workshop

%\PassOptionsToPackage{numbers}{natbib}

%\usepackage{neurips_2019}
\usepackage{iclr2020_conference,times}

@@

%\iclrfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}
\title{Zero-shot task adaptation by homoiconic meta-mapping}
\author{%
Andrew K. Lampinen\\
Department of Psychology\\
Stanford University\\
\texttt{lampinen@stanford.edu}\\
\And
James L. McClelland\\
Department of Psychology\\
Stanford University\\
\texttt{mcclelland@stanford.edu}\\
}
\date{}
\maketitle

@@

%\vspace{-0.75em} % section
\section{Introduction}
%\vspace{-0.5em} % section
Humans are able to use and reuse knowledge more flexibly than most deep learning models can \citep{Lake2016, Marcus2018}. The problem of rapid learning has been partially addressed by meta-learning systems \citep[see also section \ref{sec_discussion}]{Santoro2016, Finn2017a, Finn2018, Stadie2018, Botvinick2019}. However, humans can use our knowledge of a task to flexibly adapt when the task changes. In particular, we can often perform an altered task zero-shot, that is, without seeing any data at all. For example, once we learn to play a game, we can immediately switch to playing in order to lose, and can perform reasonably on our first attempt. \par
One fundamental reason for this is that humans are aware of what we are trying to compute and why. This allows us to adapt our task representations to perform a new task zero-shot. By contrast, most deep learning models do not explicitly represent tasks at all. Those that do (such as some meta-learning models) generally do not represent relationships between tasks, and so cannot adapt to a new task without data from the new task. To address this, it is necessary to represent tasks in a transformable space. This can grant the ability to rapidly adapt behavior to a new task. \par
In this paper, we propose a new class of tasks based on this idea: meta-mappings, i.e. mappings between tasks (see below). As noted above, this type of transfer is easily accessible to humans \citep{Lake2016}, but is generally inaccessible to deep-learning models. To address this challenge, we propose using architectures which essentially take a functional perspective on meta-learning, and exploit the idea of homoiconicity. (A homoiconic programming language is one in which programs in the language can be manipulated by programs in the language, just as data can.) By treating both data and task behaviors as functions, we can conceptually think of both data \emph{and} learned task behaviors as transformable. This yields the ability to not only learn to solve new tasks, but to learn how to transform these solutions in response to changing task demands. We demonstrate that our architectures can flexibly remap their behavior to address the meta-mapping challenge. By allowing the network to recursively treat its task representations as data points, and transform them to produce new task representations, our approach is able to achieve this flexibility parsimoniously. We suggest that approaches like ours will be key to building more intelligent and flexible deep learning systems. \par

%\vspace{-0.25em}
\section{Meta-mapping}
%\vspace{-0.5em} % section
%Can cut "the task"s in go sentence if needed
We propose the meta-mapping challenge. We define a meta-mapping as a task, or mapping, that takes a task as an input, output, or both. These include mapping from tasks to language (explaining), mapping from language to tasks (following instructions), and mapping from tasks to tasks (adapting behavior). While the first two categories have been partially addressed in prior work \citep[e.g.][]{Hermann2017, Co-Reyes2019}, the latter is more novel. (We discuss the relationship between our work and prior work in section \ref{sec_discussion}.) This adaptation can be cued in several ways, including examples of the mapping (after winning and losing at poker, try to lose at blackjack) or natural-language instructions (``try to lose at blackjack''). \par
We argue that task-to-task meta-mappings are a useful way to think about human-like flexibility, because a great deal of our rapid adaptation is from a task to some variation on that task. For example, the task of playing go on a large board is closely related to the task of playing go on a small board. Humans can exploit this to immediately play well on a different board, but deep learning models generally have no way to achieve this. We can also adapt in much deeper ways, for example fundamentally altering our value function on a task, such as trying to lose, or trying to achieve some orthogonal goal. While meta-learning systems can rapidly learn a new task from a distribution of tasks they have experience with, this does not fully capture human flexibility. Given appropriate conditioning (see below), our architecture can use meta-mappings to adapt to substantial task alterations zero-shot, that is, without seeing a single example from the new task \citep{Lake2016}. Achieving this flexibility to meta-map to new tasks will be an important step toward more general intelligence -- intelligence that is not limited to precisely the training tasks it has seen. \par

%\vspace{-0.25em}
\section{Homoiconic meta-mapping (HoMM) architecture}
%\vspace{-0.5em} % section
To address these challenges, we propose HoMM architectures, composed of two components: 
\vspace{-0.5em}
\begin{enumerate} \setlength \itemsep{0em}
\item Input/output systems: domain specific encoders and decoders (vision, language, etc.) that map into a shared embedding space $Z$.
\item A meta-learning system that a) learns to embed tasks into the shared embedding space $Z$, b) learns to use these task embeddings to perform task-appropriate behavior, c) learns to embed meta-mappings into the same space, and d) learns to use these meta-mapping embeddings to transform basic task embeddings in a meta-mapping appropriate way.\end{enumerate}
\vspace{-0.5em}
\looseness=-1
These architectures are homoiconic because they have a completely shared $Z$ for individual data points, tasks, and meta-mappings. Why is this useful? The primary advantage is that it parsimoniously allows for arbitrary mappings between these entities. In addition to basic tasks, the system can learn to perform meta-mappings to follow instructions or change behavior. That is, it can transform task representations using the same components it uses to transform basic data points. (See also appendix \ref{app_lesion_results_shared_z}.) \par
Without training on meta-mappings, of course, the system will not be able to execute them well. However, as we will show, if it is trained on a broad enough set of such mappings, it will be able to generalize to new instances drawn from the same meta-mapping distribution. For instances that fall outside its data distribution, or for optimal performance, it may require some retraining, however. This reflects the structure of human behavior -- we are able to adapt rapidly when new knowledge is relatively consistent with our prior knowledge, but learning an entirely new paradigm (such as calculus for a new student) can be quite slow \citep[cf.][]{Kumaran2016, Botvinick2019}. \par 
\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.8, every node/.style={scale=0.8}]
%% basic meta learning
\begin{scope}[shift={(0.4, 0.4)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}
\begin{scope}[shift={(0.2, 0.2)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}

@@

%\subsection{Boolean functions}
%As a proof of concept, we first evaluated the system on a simple task of computing boolean functions on boolean inputs. Specifically, we can imagine mapping from binary-valued input vectors to a single binary output representing the evaluation of some logical proposition on the input. We train the system to figure out how to compute the predicate from seeing a subset of the (input, output) pairs. \par   
%We can then train the system to do various meta-mappings, for example identifying certain types of functions (like XOR) or negating predicates. \par
%
\section{Learning multivariate polynomials} \label{sec_poly}
\begin{figure}[t]
\centering
\begin{subfigure}[t]{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/poly/basic_results.png}
\caption{The polynomials domain, section \ref{sec_poly}.}
\label{poly_basic_results}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/basic_meta_learning.png}
\caption{The cards domain, section \ref{sec_cards}.}
\label{cards_basic_results}
\end{subfigure}%
\caption{The HoMM system succeeds at basic meta-learning, which is a necessary prerequisite for meta-mappings. (\subref{poly_basic_results}) The polynomials domain, section \ref{sec_poly}. The system successfully generalizes to held out polynomials. The solid line indicates optimal performance; the dashed line indicates untrained model performance. (\subref{cards_basic_results}) The card games domain, section \ref{sec_cards}. The system successfully generalizes to held out games, both when trained on a random sample of half the tasks, or when a targeted subset is held out. The gray dashed line indicates chance performance, while the solid lines are optimal performance. The orange dashed lines shows performance on held-out tasks of playing the strategy from the most correlated trained task. The fact that the system generally exceeds this difficult baseline shows a deeper form of generalization than just memorizing strategies and picking the closest. Error-bars throughout are bootstrap 95\%-confidence intervals, numerical values for plots can be found in appendix \ref{app_numerical_results}.}
\end{figure}

@@

%\vspace{-0.5em} % section

@@

%\vspace{-0.7em}

@@

%\vspace{-0.7em}

@@

%\vspace{-0.5em}

@@

% \vspace{-1em}

>>>>

\PassOptionsToPackage{numbers}{natbib}

\usepackage[final]{neurips_2019}
%\usepackage{iclr2020_conference,times}

@@

%\iclrpreprintcopy
%\iclrfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}
\title{Zero-shot task adaptation by homoiconic meta-mapping}
\author{%
Andrew K. Lampinen\\
Department of Psychology\\
Stanford University\\
\texttt{lampinen@stanford.edu}\\
\And
James L. McClelland\\
Department of Psychology\\
Stanford University\\
\texttt{mcclelland@stanford.edu}\\
}
\date{}
\maketitle

@@

\vspace{-0.75em} % section
\section{Introduction}
\vspace{-0.5em} % section
Humans are able to use and reuse knowledge more flexibly than most deep learning models can \citep{Lake2016, Marcus2018}. The problem of rapid learning has been partially addressed by meta-learning systems \citep[see also section \ref{sec_discussion}]{Santoro2016, Finn2017a, Finn2018, Stadie2018, Botvinick2019}. However, humans can use our knowledge of a task to flexibly adapt when the task changes. In particular, we can often perform an altered task zero-shot, that is, without seeing any data at all. For example, once we learn to play a game, we can immediately switch to playing in order to lose, and can perform reasonably on our first attempt. \par
One fundamental reason for this is that humans are aware of what we are trying to compute and why. This allows us to adapt our task representations to perform a new task zero-shot. By contrast, most deep learning models do not explicitly represent tasks at all. Those that do (such as some meta-learning models) generally do not represent relationships between tasks, and so cannot adapt to a new task without data from the new task. To address this, it is necessary to represent tasks in a transformable space. This can grant the ability to rapidly adapt behavior to a new task. \par
In this paper, we propose a new class of tasks based on this idea: meta-mappings, i.e. mappings between tasks (see below). As noted above, this type of transfer is easily accessible to humans \citep{Lake2016}, but is generally inaccessible to deep-learning models. To address this challenge, we propose using architectures which essentially take a functional perspective on meta-learning, and exploit the idea of homoiconicity. (A homoiconic programming language is one in which programs in the language can be manipulated by programs in the language, just as data can.) By treating both data and task behaviors as functions, we can conceptually think of both data \emph{and} learned task behaviors as transformable. This yields the ability to not only learn to solve new tasks, but to learn how to transform these solutions in response to changing task demands. We demonstrate that our architectures can flexibly remap their behavior to address the meta-mapping challenge. By allowing the network to recursively treat its task representations as data points, and transform them to produce new task representations, our approach is able to achieve this flexibility parsimoniously. We suggest that approaches like ours will be key to building more intelligent and flexible deep learning systems. \par

\vspace{-0.25em}
\section{Meta-mapping}
\vspace{-0.5em} % section
%Can cut "the task"s in go sentence if needed
We propose the meta-mapping challenge. We define a meta-mapping as a task, or mapping, that takes a task as an input, output, or both. These include mapping from tasks to language (explaining), mapping from language to tasks (following instructions), and mapping from tasks to tasks (adapting behavior). While the first two categories have been partially addressed in prior work \citep[e.g.][]{Hermann2017, Co-Reyes2019}, the latter is more novel. (We discuss the relationship between our work and prior work in section \ref{sec_discussion}.) This adaptation can be cued in several ways, including examples of the mapping (after winning and losing at poker, try to lose at blackjack) or natural-language instructions (``try to lose at blackjack''). \par
We argue that task-to-task meta-mappings are a useful way to think about human-like flexibility, because a great deal of our rapid adaptation is from a task to some variation on that task. For example, the task of playing go on a large board is closely related to the task of playing go on a small board. Humans can exploit this to immediately play well on a different board, but deep learning models generally have no way to achieve this. We can also adapt in much deeper ways, for example fundamentally altering our value function on a task, such as trying to lose, or trying to achieve some orthogonal goal. While meta-learning systems can rapidly learn a new task from a distribution of tasks they have experience with, this does not fully capture human flexibility. Given appropriate conditioning (see below), our architecture can use meta-mappings to adapt to substantial task alterations zero-shot, that is, without seeing a single example from the new task \citep{Lake2016}. Achieving this flexibility to meta-map to new tasks will be an important step toward more general intelligence -- intelligence that is not limited to precisely the training tasks it has seen. \par

\vspace{-0.25em}
\section{Homoiconic meta-mapping (HoMM) architecture}
\vspace{-0.5em} % section
To address these challenges, we propose HoMM architectures, composed of two components: 
\vspace{-0.5em}
\begin{enumerate} \setlength \itemsep{0em}
\item Input/output systems: domain specific encoders and decoders (vision, language, etc.) that map into a shared embedding space $Z$.
\item A meta-learning system that a) learns to embed tasks into the shared embedding space $Z$, b) learns to use these task embeddings to perform task-appropriate behavior, c) learns to embed meta-mappings into the same space, and d) learns to use these meta-mapping embeddings to transform basic task embeddings in a meta-mapping appropriate way.\end{enumerate}
\vspace{-0.5em}
\looseness=-1
These architectures are homoiconic because they have a completely shared $Z$ for individual data points, tasks, and meta-mappings. Why is this useful? The primary advantage is that it parsimoniously allows for arbitrary mappings between these entities. In addition to basic tasks, the system can learn to perform meta-mappings to follow instructions or change behavior. That is, it can transform task representations using the same components it uses to transform basic data points. (See also appendix \ref{app_lesion_results_shared_z}.) \par
Without training on meta-mappings, of course, the system will not be able to execute them well. However, as we will show, if it is trained on a broad enough set of such mappings, it will be able to generalize to new instances drawn from the same meta-mapping distribution. For instances that fall outside its data distribution, or for optimal performance, it may require some retraining, however. This reflects the structure of human behavior -- we are able to adapt rapidly when new knowledge is relatively consistent with our prior knowledge, but learning an entirely new paradigm (such as calculus for a new student) can be quite slow \citep[cf.][]{Kumaran2016, Botvinick2019}. \par 
\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.8, every node/.style={scale=0.8}]
%% basic meta learning
\begin{scope}[shift={(0.4, 0.4)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}
\begin{scope}[shift={(0.2, 0.2)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}

@@

%\subsection{Boolean functions}
%As a proof of concept, we first evaluated the system on a simple task of computing boolean functions on boolean inputs. Specifically, we can imagine mapping from binary-valued input vectors to a single binary output representing the evaluation of some logical proposition on the input. We train the system to figure out how to compute the predicate from seeing a subset of the (input, output) pairs. \par   
%We can then train the system to do various meta-mappings, for example identifying certain types of functions (like XOR) or negating predicates. \par
%
\section{Learning multivariate polynomials} \label{sec_poly}
\vspace{-0.8em}
\begin{figure}[t]
\centering
\begin{subfigure}[t]{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/poly/basic_results.png}
\caption{The polynomials domain, section \ref{sec_poly}.}
\label{poly_basic_results}
\end{subfigure}%
\begin{subfigure}[t]{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/basic_meta_learning.png}
\caption{The cards domain, section \ref{sec_cards}.}
\label{cards_basic_results}
\end{subfigure}%
\caption{The HoMM system succeeds at basic meta-learning, which is a necessary prerequisite for meta-mappings. (\subref{poly_basic_results}) The polynomials domain, section \ref{sec_poly}. The system successfully generalizes to held out polynomials. The solid line indicates optimal performance; the dashed line indicates untrained model performance. (\subref{cards_basic_results}) The card games domain, section \ref{sec_cards}. The system successfully generalizes to held out games, both when trained on a random sample of half the tasks, or when a targeted subset is held out. The gray dashed line indicates chance performance, while the solid lines are optimal performance. The orange dashed lines shows performance on held-out tasks of playing the strategy from the most correlated trained task. The system generally exceeds this difficult baseline, thus showing deeper generalization than just memorizing strategies and picking the closest. Error-bars are bootstrap 95\%-CIs, numerical values for plots can be found in appendix \ref{app_numerical_results}.}
\vspace{-0.5em}
\end{figure}

@@

\vspace{-0.5em} % section

@@

\vspace{-1em}

@@

\vspace{-0.7em}

@@

@@

\vspace{-1em}
